Accelerating Biomolecular Nuclear Magnetic Resonance Assignment with A*

### 2014-03-26 paper/tex/Paper.tex
Added results section to paper

A prerequisite of the assignment process is data collection using experiments that can provide connectivities between neighboring residues. For example, the HNCACB experiment identifies the chemical shifts corresponding to the $C_\alpha$ and $C_{\beta}$ nuclei of one residue (residue i), as well as the $C_\alpha$ and $C_{\beta}$ chemical shifts of the immediately preceding residue (residue $i – 1$). In this experiment, the values corresponding to residue i can usually be distinguished from the ($i -1$) values by their higher intensity. However, ambiguities can arise if the intensities are comparable or if the chemical shift values are overlapping. These ambiguities can be resolved by using additional experiments, such as CBCA(CO) NH, which yields the chemical shifts of the preceding residue only. Using all inter-residue connectivities, a chain of correlations through the protein backbone chemical shifts can be established.  This pattern of sequentially linked chemical shift values reflects the sequential linear arrangement of the individual residues in the protein sequence. The pattern is then matched with the protein sequence, by using the fact that certain residues have characteristic $C_\alpha$ and $C_{\beta}$ chemical shift ranges which uniquely identify them. Thus, each measured chemical shift is assigned to a protein residue, and this information can then be used to infer structural information about the biomolecule.

@@

\subsection{Algorithm} % (fold)
\label{sub:algorithm}
Our algorithm begins by reading in an expected amino acid order and the NMR carbon shift data from a data file. It then changes the amino acids to their expected $C_\alpha$ and $C_{\beta}$ values and stores them in an list that represents the protein chain. Next, the carbon shift values for a single amino acid are placed in a tile. A tile contains the $C_\alpha$ and $C_{\beta}$ values for the amino acid residue $i$ and residue $i-1$. Next, missing data is accounted for. If there are less tiles than the number of amino acids in the protein chain, placeholder tiles are generated to make up the difference. Before the search for the best solution begins, the tiles are grouped by the type of amino acids they could represent to accelerate the assignment process. Tiles that do not match a group are grouped with the placeholder tiles.

%start search
To start the search, tiles in the placeholder group and tiles of the same amino acid group as the first amino acid in the protein chain are placed in the first position in a node. The tile is then compared to the first amino acid in the protein chain and an cost is generated. The cost is the sum of the cost of placing any previous tiles in the node, how well the current tile matched the corresponding amino acid in the protein chain and how well the tile's residue $i-1$ values match the residue $i$ values of the tile placed before it. All of the generated nodes are place in a frontier array.

%continue assignment
The node in the frontier with the lowest cost is removed from the frontier and used to generate child nodes. To generate a child node, a tile is added to the end of the node. A child node is generated for all unplaced tiles that are either in the placeholder group or in the group matching the corresponding amino acid in the protein chain. A cost of placing the tile is then generated.  The child nodes are then added to the frontier, and the process of generating child nodes is repeated.

@@

>>>>

A prerequisite of the assignment process is data collection using experiments that can provide connectivities between neighboring residues. For example, the HNCACB experiment identifies the chemical shifts corresponding to the $C_\alpha$ and $C_{\beta}$ nuclei of one residue (residue $i$), as well as the $C_\alpha$ and $C_{\beta}$ chemical shifts of the immediately preceding residue (residue $i – 1$). In this experiment, the values corresponding to residue $i$ can usually be distinguished from the ($i -1$) values by their higher intensity. However, ambiguities can arise if the intensities are comparable or if the chemical shift values are overlapping. These ambiguities can be resolved by using additional experiments, such as CBCA(CO) NH, which yields the chemical shifts of the preceding residue only. Using all inter-residue connectivities, a chain of correlations through the protein backbone chemical shifts can be established.  This pattern of sequentially linked chemical shift values reflects the sequential linear arrangement of the individual residues in the protein sequence. The pattern is then matched with the protein sequence, by using the fact that certain residues have characteristic $C_\alpha$ and $C_{\beta}$ chemical shift ranges which uniquely identify them. Thus, each measured chemical shift is assigned to a protein residue, and this information can then be used to infer structural information about the biomolecule.

@@

\subsection{Algorithm} % (fold)
\label{sub:algorithm}
Our algorithm begins by reading in an expected amino acid order and the NMR carbon shift data from a data file. It then changes the amino acids to their expected $C_\alpha$ and $C_{\beta}$ values and stores them in a list that represents the protein chain. Next, the carbon shift values for a single amino acid are placed in a tile. A tile contains the $C_\alpha$ and $C_{\beta}$ values for the amino acid residue $i$ and residue $i-1$. Next, missing data is accounted for. If there are fewer tiles than the number of amino acids in the protein chain, placeholder tiles are generated to make up the difference. Before the search for the best solution begins, the tiles are grouped by the types of amino acids they could represent to accelerate the assignment process. Tiles that do not match a group are grouped with the placeholder tiles.

%start search
To start the search, tiles in the placeholder group and tiles of the same amino acid group as the first amino acid in the protein chain are placed in the first position in a node. The tile is then compared to the first amino acid in the protein chain and a cost is generated. The cost is the sum of the cost of placing any previous tiles in the node, how well the current tile matched the corresponding amino acid in the protein chain and how well the tile's residue $i-1$ values match the residue $i$ values of the tile placed before it. All of the generated nodes are placed in a frontier array.

%continue assignment
The node in the frontier with the lowest cost is removed from the frontier and used to generate child nodes. To generate a child node, a tile is added to the end of the node. A child node is generated for all unplaced tiles that are either in the placeholder group or in the group matching the corresponding amino acid in the protein chain. The cost of placing the tile is then generated.  The child nodes are then added to the frontier, and the process of generating child nodes is repeated.

@@

\section{Conclusions}
\label{sec:conclusions}

\subsection{Results}
\label{sub:results}
Our algorithm has been tested with real NMR datasets of varying size. When compared to hand-ordered data, the algorithm produces the same, optimal ordering of amino acids. Initially, we focused our research on accurate assignment at the cost of speed. Thus, at the start of our research we could assign only smaller datasets up to ten to fifteen amino acids long. As we shifted our focus to optimizing the algorithm's speed while maintaining the accuracy we managed to achieve we improved runtime to correctly assign up to forty to fifty amino acids. In future research, we plan to test with larger datasets, maximizing speed and efficiency to the point that our algorithm can assign full protein sequences. We also plan to look into testing datasets with more than a few missing datapoints to verify that our algorithm correctly places amino acids when given less-perfect data.
